\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-29 chapters/wcble.tex
Layout chapter wcble
Second layout run

\chapter{\Acs{erp} latency estimation and alignment}
\label{sec:wcble}
\epigraph{``The sands of time cannot be stopped.''}{Brom in \emph{Eragon --
Christopher Paolini}}
\noindent\emph{\Cref{sec:wcble/literature/contrib},
\crefrange{sec:wcble/methods/cble}{sec:wcble/methods/wcble} (including
\cref{fig:diagram-train,fig:diagram-eval}), and the third-to-last paragraph of
\cref{sec:wcble/conclusion} were published as part of~\textcite{VanDenKerchove2024}.}

@@

\Ac{eeg}-based visual oddball \acp{bci} establish communication with paralyzed
individuals by decoding \acp{erp} obtained from time-modulated flashing
stimuli (see~\cref{fig:bci/loop}).
These \acp{erp} consist of multiple \ac{erp}-components, peaks and troughs in
\ac{erp} waveform (see \cref{fig:bci/erp}).
These components each have their own distinct amplitude and latency.
\Ac{erp} component latency can reflect a wide range of cognitive,
neurophysiological, and clinical properties, that can be confounding factors in
the development of \acp{bci}~\cite{Luck2014}.
Latency is often assessed as the timing of an \ac{erp} component peak, obtained
after averaging over trials.
Latencies of non-averaged, single-trial \acp{erp} should capture even more
relevant information, yet they are very hard to measure due to the low \ac{snr} of
single-trial \acp{erp}.
Within-session variability of \ac{erp} latency, called \ac{erp} latency
\emph{jitter}, has been shown to influence \ac{bci}
accuracy~\cite{Thompson2012}.
This effect is especially prominent when bringing \ac{bci} development from the
lab to the clinical setting.
For instance, \textcite{Zisk2021} showed that \ac{erp} jitter was significantly higher
in 6 \ac{als} patients (1350 ms$^2$) compared to neurotypical controls
(553 ms$^2$).
\textcite{Arico2014} showed that P3 latency jitter negatively correlated
($\rho=0.17$) with \ac{bci} accuracy in a covert GeoSpell interface.
Here the user does not gaze directly at the targets, such as is
the case for individuals with \ac{sspgi}.
Therefore, properly accounting for \ac{erp} latency jitter can contribute to
the development of gaze-independent visual \acp{bci}.

@@

\begin{figure}
  \centering
  \input{figures/wcble/smearing.tikz.tex}
  \caption[The smearing effect.]{The smearing effect. When obtaining an \ac{erp} template through
  averaging, high jitter causes a decrease in peak amplitude and a deformation
  of the shape of the waveform.}
  \label{fig:wcble/smearing}
\end{figure}

@@

The \ac{wcble} algorithm is presented in \cref{sec:wcble/app/wcble-alg}.
Its training phase is visualized in~\cref{fig:diagram-train}.
The initial training epochs $\{\mat{X}_n^{(1)}\}_{n=1}^N$ are set to $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N$.
At every iteration, classifier $\mathcal{C}$ is trained like in \ac{cble}:
\begin{equation}
  \theta^{(i)} =
  \mathrm{train}_\mathcal{C}(\{\mat{X}^{(i-1)}_n[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
\end{equation}
Next, latency $s_{l_n}^{(i)}$ is determined for every epoch $\mat{X}^{(i)}$ corresponding
to its class label $l_n$ using \cref{eq:prob-latency}.
Finally, the training epochs $\mat{X}^{(i+1)}$ for the next iteration are determined by aligning
each original training epoch to the latency $s_{l_n}^{(i)}$ corresponding to its respective class
label.
\begin{equation}
  \mat{X}^{(i+1)}_n = \mathrm{align}(\mat{X}_n^\mathrm{train}, s_{l_n}^{(i)}) \quad \forall n=1,\ldots,N
\end{equation}
Aligning is performed by shifting and zero-padding the signal to the right if
the latency is negative relative to the time window onset, and to the left if
positive, by the difference between the latency and the window onset.
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/figure1.pdf}
}{%
  \caption[Schematic representation of the \acs{wcble} training phase.]{%
    Schematic representation of the \acf{wcble} training phase.
    (1) The first-stage spatiotemporal binary classifier is trained on a set of
    epochs.
    (2) It is then applied to time-shifted copies of the epochs to obtain scores
    and class probabilities over time.
    (3) The medians of these probability distributions are assumed as the class
    latencies.
    (4) The epochs are aligned to their corresponding class latencies by
    shifting in time such that all latencies fall at the same moment.
    (5) The spatiotemporal classifier is then retrained on the aligned epochs
    for the next iteration.
    (6) After the iterative process halts, the scores and latencies obtained
    from the last iteration are used to train the second-stage
    classifier.
  }
  \label{fig:diagram-train}
}
The process halts after a fixed number of iterations or when the estimated set
of latencies has been encountered before, indicating it ended up in a loop.
In the end, the procedure should result in enhanced classifier parameters $\theta^*$,
closer to those when there would be no jitter between epochs.
Note that using the median approach for robust latency estimation results in a
smoother yet longer convergence process compared to the maximum/minimum approach.
We can then apply the classifier with enhanced parameters $\theta^*$ in a \ac{cble}
manner to unseen epochs as illustrated in \cref{fig:diagram-eval} to obtain a
vector of scores over time as in \cref{sec:wcble/methods/cble} and the estimated latencies as
in~\cref{sec:robust-latency}.
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/figure2.pdf}
}{%
  \caption[Schematic representation of the \acs{wcble} test phase.]{%
    Schematic representation of the \acf{wcble}. (1) The
	  first-stage spatiotemporal binary classifier obtained from the training phase is
	  applied to time-shifted copies of the epochs to obtain scores and class probabilities
	  over time. (2) The medians of these probability distributions are assumed
	  as the new class latencies. (3) The scores and class latencies are input to the
    trained second-stage classifier, which predicts the label of the epochs.
  }
  \label{fig:diagram-eval}
}

@@

A sample of the simulated evoked data is displayed in \cref{fig:wcble/sim-sine}.
\fullpagefig{%
  \inputpgf{figures/wcble}{simulated-sine.pgf}
}{%
  \caption[Simulated target and non-target evoked data.]{%
    Simulated target (bottom half of epochs) and non-target (top half of epochs) evoked data
    and their average (red: target, yellow: non-target, shaded area: standard
    deviation).
    Targets contain a sine wave pulse jittered by $\sigma$. 32 \ac{eeg} channels with pink
    spatiotemporal noise were simulated using a forward head model, and channel C3 is
    plotted here.
  }
  \label{fig:wcble/sim-sine}
}

@@

\begin{figure}
  \sffamily\sansmath
  \input{figures/wcble/fig_latency.pgf}
  \caption[Synthetic latency estimation results]{Correctness of estimated
  latency for varying jitter ($\sigma$) and \ac{snr} introduced in synthetic evoked target
  data, measured as Pearson correlation coefficient $\rho$ between
  estimated and ground truth latencies.
  \Ac{wcble} estimates latencies more accurately in the presence of noise and
  high jitter.}
  \label{fig:wcble/results/latency}
\end{figure}

@@

\begin{figure}
  \sffamily\sansmath
  \input{figures/wcble/fig_accuracy.pgf}
  \caption[Synthetic decoding results]{Decoding performance, measured as binary
  classification accuracy, for \ac{cble}, \ac{wcble}, and their first stage
  classifier \ac{tlda} for varying introduced jitter ($\sigma$) and \ac{snr} in
  synthetic evoked target and non-target data.
  \Ac{wcble} decoding performance is robust to more noise and jitter than the other methods.}
  \label{fig:wcble/results/accuracy}
\end{figure}

@@

@@

\clearpage
\section{\Acs{wcble} algorithm}
\label{sec:wcble/app/wcble-alg}
\begin{algorithm}[h]
	\textsc{Train}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N, \mathbf{l^\mathrm{train}},
		\mathcal{C}(\cdot,f,\Pr), s_1, s_2$
	\begin{algorithmic}[1]
		\State $\mat{X}'_n \gets \mat{X}^\mathrm{train}_n \quad \forall \quad n=1\ldots
    N$
      \algorithmiccomment{Train stage 1}
		\Repeat
		\State $\theta^* \gets \mathrm{train}_\mathcal{C}(\{\mat{X}'_n[:,s_1:s_2]\}_0^{N-1},\mathbf{l^\mathrm{train}})$
		\For{$n=1\ldots N$}
    \State  $s_n \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n',\theta^*,l_n)\right]$
		\State $\mat{X}'_n \gets \mathrm{align}(\mat{X}^\mathrm{train}_n, s^*_n)$
		\EndFor
    \Until{convergence or maximum iterations reached}
    \For{$n=1\ldots N$}\algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{n,s}^\mathrm{train} \gets
      f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta^*)$
        \EndFor
        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\mathrm{target})\right]$
		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\textrm{non-target})\right]$
    \EndFor

>>>>

\chapter{\Acs{erp} latency estimation and alignment}
\label{sec:wcble}
\epigraph{``The sands of time cannot be stopped.''}{Brom in \emph{Eragon --
Christopher Paolini}}
\publicationnote{\Cref{sec:wcble/literature/contrib},
\crefrange{sec:wcble/methods/cble}{sec:wcble/methods/wcble} (including
\cref{fig:diagram-train,fig:diagram-eval}), and the third-to-last paragraph of
\cref{sec:wcble/conclusion} were published as part of \textcite{VanDenKerchove2024}.}

@@

\Ac{eeg}-based visual oddball \acp{bci} establish communication with paralyzed
individuals by decoding \acp{erp} obtained from time-modulated flashing
stimuli (see~\cref{fig:bci/loop}).
These \acp{erp} consist of multiple \ac{erp}-components, peaks and troughs in
\ac{erp} waveform (see \cref{fig:bci/erp}).
These components each have their own distinct amplitude and latency.
\Ac{erp} component latency can reflect a wide range of cognitive,
neurophysiological, and clinical properties, that can be confounding factors in
the development of \acp{bci}~\cite{Luck2014}.
Latency is often assessed as the timing of an \ac{erp} co, **kwargsmponent peak, obtained
after averaging over trials.
Latencies of non-averaged, single-trial \acp{erp} should capture even more
relevant information, yet they are very hard to measure due to the low \ac{snr} of
single-trial \acp{erp}.
Within-session variability of \ac{erp} latency, called \ac{erp} latency
\emph{jitter}, has been shown to influence \ac{bci}
accuracy~\cite{Thompson2012}.
This effect is especially prominent when bringing \ac{bci} development from the
lab to the clinical setting.
For instance, \textcite{Zisk2021} showed that \ac{erp} jitter was significantly higher
in 6 \ac{als} patients (1350 ms$^2$) compared to neurotypical controls
(553 ms$^2$).
\textcite{Arico2014} showed that P3 latency jitter negatively correlated
($\rho=0.17$) with \ac{bci} accuracy in a covert GeoSpell interface.
Here the user does not gaze directly at the targets, such as is
the case for individuals with \ac{sspgi}.
Therefore, properly accounting for \ac{erp} latency jitter can contribute to
the development of gaze-independent visual \acp{bci}.

@@

\begin{figure}[t]
  \centering
  \input{figures/wcble/smearing.tikz.tex}
  \caption[The smearing effect.]{The smearing effect. When obtaining an \ac{erp} template through
  averaging, high jitter causes a decrease in peak amplitude and a deformation
  of the shape of the waveform.}
  \label{fig:wcble/smearing}
\end{figure}

@@

The \ac{wcble} algorithm is presented in \cref{sec:wcble/app/wcble-alg}.
Its training phase is visualized in~\cref{fig:diagram-train}.
The initial training epochs $\{\mat{X}_n^{(1)}\}_{n=1}^N$ are set to $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N$.
At every iteration, classifier $\mathcal{C}$ is trained like in \ac{cble}:
\begin{equation}
  \theta^{(i)} =
  \mathrm{train}_\mathcal{C}(\{\mat{X}^{(i-1)}_n[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
\end{equation}
Next, latency $s_{l_n}^{(i)}$ is determined for every epoch $\mat{X}^{(i)}$ corresponding
to its class label $l_n$ using \cref{eq:prob-latency}.
Finally, the training epochs $\mat{X}^{(i+1)}$ for the next iteration are determined by aligning
each original training epoch to the latency $s_{l_n}^{(i)}$ corresponding to its respective class
label.
\begin{equation}
  \mat{X}^{(i+1)}_n = \mathrm{align}(\mat{X}_n^\mathrm{train}, s_{l_n}^{(i)}) \quad \forall n=1,\ldots,N
\end{equation}
Aligning is performed by shifting and zero-padding the signal to the right if
the latency is negative relative to the time window onset, and to the left if
positive, by the difference between the latency and the window onset.
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/figure1.pdf}
}{%
  \caption[Schematic representation of the \acs{wcble} training phase.]{%
    The \acf{wcble} training phase.
    (1) The first-stage classifier is trained on a set of
    epochs.
    (2) It is then applied to time-shifted copies of the epochs, yielding scores
    and class probabilities over time.
    (3) The medians of the probability distributions are assumed as the class
    latencies.
    (4) The epochs are aligned by shifting in time based on their corresponding class latencies by
    shifting in time.
    (5) The spatiotemporal classifier is then retrained on the aligned epochs
    for the next iteration.
    (6) On convergence, the scores and latencies obtained
    from the last iteration are used to train the second-stage
    classifier.
  }
  \label{fig:diagram-train}
}
The process halts after a fixed number of iterations or when the estimated set
of latencies has been encountered before, indicating it ended up in a loop.
In the end, the procedure should result in enhanced classifier parameters $\theta^*$,
closer to those when there would be no jitter between epochs.
Note that using the median approach for robust latency estimation results in a
smoother yet longer convergence process compared to the maximum/minimum approach.
We can then apply the classifier with enhanced parameters $\theta^*$ in a \ac{cble}
manner to unseen epochs as illustrated in \cref{fig:diagram-eval} to obtain a
vector of scores over time as in \cref{sec:wcble/methods/cble} and the estimated latencies as
in~\cref{sec:robust-latency}.
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/figure2.pdf}
}{%
  \caption[Schematic representation of the \acs{wcble} test phase.]{%
    The \acf{wcble} test phase. (1) The
	  first-stage classifier obtained from the training phase is
	  applied to time-shifted copies of the epochs, yielding scores and class probabilities
	  over time. (2) The medians of these probability distributions are assumed
	  as the new class latencies. (3) The scores and class latencies are input to the
    trained second-stage classifier, which predicts the label of the epochs.
  }
  \label{fig:diagram-eval}
}

@@

A sample of the simulated evoked data is displayed in \cref{fig:wcble/sim-sine}.
\fullpagefig[anchor.north]{%
  \hspace{1cm}\inputpgf{figures/wcble}{simulated-sine.pgf}
}{%
  \caption[Simulated target and non-target evoked data.]{%
    Simulated target (bottom half of epochs) and non-target (top half of epochs) evoked data
    and their average (red: target, yellow: non-target, shaded area: standard
    deviation).
    Targets contain a sine wave pulse jittered by $\sigma$. 32 \ac{eeg} channels with pink
    spatiotemporal noise were simulated using a forward head model, and channel C3 is
    plotted here.
  }
  \label{fig:wcble/sim-sine}
}

@@

@@

@@

\begin{figure}[t!]
  \input{figures/wcble/fig_latency.pgf}
  \caption[Synthetic latency estimation results]{Correctness of estimated
  latency for varying jitter ($\sigma$) and \ac{snr} introduced in synthetic evoked target
  data, measured as Pearson correlation coefficient $\rho$ between
  estimated and ground truth latencies.
  \Ac{wcble} estimates latencies more accurately in the presence of noise and
  high jitter.}
  \label{fig:wcble/results/latency}
\end{figure}

\begin{figure}[t!]
  \sffamily\sansmath
  \input{figures/wcble/fig_accuracy.pgf}
  \caption[Synthetic decoding results]{Decoding performance, measured as binary
  classification accuracy, for \ac{cble}, \ac{wcble}, and their first stage
  classifier \ac{tlda} for varying introduced jitter ($\sigma$) and \ac{snr} in
  synthetic evoked target and non-target data.
  \Ac{wcble} decoding performance is robust to more noise and jitter than the other methods.}
  \label{fig:wcble/results/accuracy}
\end{figure}

@@

\clearpage
\section{\Acs{wcble} algorithm}
\label{sec:wcble/app/wcble-alg}
\begin{algorithm}[h]
  \small
	\textsc{Train}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N, \mathbf{l^\mathrm{train}},
		\mathcal{C}(\cdot,f,\Pr), s_1, s_2$
	\begin{algorithmic}[1]
		\State $\mat{X}'_n \gets \mat{X}^\mathrm{train}_n \quad \forall \quad n=1\ldots
    N$
      \algorithmiccomment{Train stage 1}
		\Repeat
		\State $\theta^* \gets \mathrm{train}_\mathcal{C}(\{\mat{X}'_n[:,s_1:s_2]\}_0^{N-1},\mathbf{l^\mathrm{train}})$
		\For{$n=1\ldots N$}
    \State  $s_n \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n',\theta^*,l_n)\right]$
		\State $\mat{X}'_n \gets \mathrm{align}(\mat{X}^\mathrm{train}_n, s^*_n)$
		\EndFor
    \Until{convergence or maximum iterations reached}
    \For{$n=1\ldots N$}\algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{n,s}^\mathrm{train} \gets
      f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta^*)$
        \EndFor
        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\mathrm{target})\right]$
		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\textrm{non-target})\right]$
    \EndFor
